---
title: "RepData_PeerAssessment1"
author: "franceshuisman"
date: "April 12, 2016"
output: 
        html_document: 
          keep_md: yes
---



## Part 1
### Reading and preprocessing the data

First load dependencies, in this case dplyr and lattice.

```{r dependencies}
library(dplyr)
library(lattice)
```

Then read in the data and convert the 'date' column to date format. This is
necessary for later processing.

```{r read-in}
d <- read.csv("repdata_data_activity/activity.csv")
d <- mutate(d, date = as.Date.factor(date))
```


## Part 2
### Steps taken per day

Group and summarise the step data by date.

```{r stepsperday1}
perday <- d %>% group_by(date) %>% summarise(steps = sum(steps, na.rm = TRUE))
```

Plot daily totals as a histogram.

```{r stepsperday2}
with(perday, hist(steps, xlab = "Number of steps", 
                  main = "Frequency of total daily step count"))
```

Calculate the mean and median of the daily totals.

```{r stepsperday3}
mean(perday$steps)
median(perday$steps)
```


## Part 3
### Average daily activity pattern

Group and summarise the step data by interval.

```{r perinterval1}
perinterval <- d %>% group_by(interval) %>% summarise(steps = mean(steps, 
                                                             na.rm = TRUE))
```

Plot a time series of the average number of steps per interval.

```{r perintervall2}
p4 <- xyplot(steps ~ interval, data = perinterval, type = "l", 
             xlab = "Interval", ylab = "Average number of steps", 
             main = "Average daily activity pattern")
print(p4)
```

Find the 5-minute interval with the maximum average number of steps.

```{r perinterval3}
with(perinterval, interval[which(steps == max(steps))])
```


## Part 4
### Impute missing values

Calculate the number of rows with NA values.

```{r impute1}
sum(is.na(d$steps))
```

Fill in the missing data. In this step, NA values in the *steps* column are
replaced with the mean number of steps for that interval.

```{r impute2}
dfixed <- d %>% group_by(interval) %>% mutate(stepsfixed = replace(steps,
                                is.na(steps), mean(steps, na.rm = TRUE)))
```

As before, group and summarise step data by date, and then plot daily totals
as a histogram.

```{r impute3}
perdayfixed <- dfixed %>% group_by(date) %>% 
        summarise(stepsfixed = sum(stepsfixed))
with(perdayfixed, hist(stepsfixed, xlab = "Number of steps", 
                  main = "Frequency of total daily step count (fixed)"))
```

Calculate the mean and median of the new daily totals.

```{r impute4}
mean(perdayfixed$stepsfixed)
median(perdayfixed$stepsfixed)
```

These numbers are identical, as any days without step data were imputed to have 
mean results, resulting in a number of counts of statistically average results. 


## Part 5
### Weekday/weekend activity patterns

Assign *weekday* and *weekend* factors based on the day of the week.

```{r daytype1}
dfixed <- dfixed %>% mutate(daytype = ifelse(weekdays(date) 
                        %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

Group and summarise data by interval and the new *daytype* factors.

```{r daytype2}
bydaytype <- dfixed %>% group_by(daytype, interval) %>% 
        summarise(steps = mean(stepsfixed))
```

Plot separate time series of the average number of steps per interval for 
weekdays and weekends.

```{r daytype3}
p8 <- xyplot(steps ~ interval | daytype, data = bydaytype, type = "l", 
             layout = c(1, 2), xlab = "Interval", 
             ylab = "Average number of steps", 
             main = "Average daily activity pattern")
print(p8)
```
